---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The activity.zip file is unzipped, the zipped csv file is read into R, and then pre-processed by converting the date to POSIXct/POSIXt format using the ymd function in the lubridate package. Note: the activity.zip file must be in the working directory.  

```{r setwd}
library(knitr)
opts_chunk$set(warning=FALSE)
setwd("F:/Documents/Education/Coursera_DataScience/RepData_PeerAssessment1")
```

```{r loadPreprocess, cache=TRUE}
library(lubridate)
unzip("./activity.zip",overwrite=TRUE)
activity <- read.csv("./activity.csv",header=TRUE,na.strings="NA",stringsAsFactors=FALSE,colClasses=c("numeric","character","numeric"))
activity$dateFactor <- as.factor(activity$date)
activity$date <- ymd(activity$date)
```

## What is mean total number of steps taken per day?
The activity dateframe is subsetted to include only those values with steps measured (i.e., non-NA values). The total steps per day are calculated and then dislayed as a histogram below, along with the mean and median total steps per day.  

```{r meanCalcs, fig.height=4}
library(ggplot2)
# subset data to only include step measurements with values (non-NA values)
activeDays <- activity[!is.na(activity$steps),]
# calculate total number of steps taken per day
totalSteps <- tapply(activeDays$steps,activeDays$dateFactor,sum)
totalSteps <- as.data.frame(totalSteps)
totalSteps$date <- as.factor(row.names(totalSteps))
names(totalSteps) <- c("totSteps","date")
# create histogram of total steps taken per day
stepHisto <- ggplot(totalSteps,aes(x=totSteps)) + geom_histogram(binwidth=1000,color="black",fill="dodgerblue4") + labs(x="Total Steps Taken Per Day",y="Frequency")
print(stepHisto)
# calculate mean and median total steps taken per day
meanTotSteps <- mean(totalSteps$totSteps,na.rm=TRUE)
meanTotSteps
medianTotSteps <- median(totalSteps$totSteps,na.rm=TRUE)
medianTotSteps
```

The mean total number of steps taken per day is `r meanTotSteps` and the median total number of steps taken per day is `r medianTotSteps`.  

## What is the average daily activity pattern?
The daily activity pattern is described as the mean number of steps taken in each 5 minute interval over the course of a day, ranging from 0 to 2355 (corresponding to 12:00am to 11:55pm).  
```{r dailyPattern, fig.height=4}
# create factor variable for the interval value
activeDays$intFactor <- factor(activeDays$interval)
# calculate mean number of steps taken in each 5 minute interval
meanStepPattern <- as.data.frame(tapply(activeDays$steps,activeDays$intFactor,mean))
meanStepPattern$interval <- as.numeric(row.names(meanStepPattern))
names(meanStepPattern) <- c("meanSteps","interval")
# determine the 5-minute interval during which, on average, the most steps are taken
maxInt <- meanStepPattern$interval[which.max(meanStepPattern$meanSteps)]
# create time series plot of mean number of steps as a function of 5-min interval
pattern <- ggplot(meanStepPattern,aes(x=interval,y=meanSteps)) + geom_line(color="dodgerblue4") + labs(x="Time Interval",y="Mean Steps",title="Mean Steps Taken as a Function of Time of Day")
print(pattern)
maxInt
```
The 5-minute time interval during which, on average, the most steps are taken is `r maxInt`.  

## Inputing missing values
```{r missingValues}
# calculate number of NA values
numNA <- sum(is.na(activity$steps))
numNA
# calculate number of days of NA values
daysNA <- numNA/288 # 12 five-minute intervals per hour = 288 five-minute intervals per day
idxNA <- which(is.na(activity$steps))
activity.fill <- activity
# calculate rounded mean number of steps taken per 5 minute interval across all days
meanStepPattern$meanStepsRound <- round(meanStepPattern$meanSteps)
# fill in NA values with the rounded mean values
for (i in 1:length(idxNA)) {
    activity.fill$steps[idxNA[i]] <- meanStepPattern$meanStepsRound[meanStepPattern$interval==activity.fill$interval[idxNA[i]]]
}
# calculate total number of steps taken per day for the filled in dataset
totalStepsFill <- tapply(activity.fill$steps,activity.fill$dateFactor,sum)
totalStepsFill <- as.data.frame(totalStepsFill)
totalStepsFill$date <- as.factor(row.names(totalStepsFill))
names(totalStepsFill) <- c("totSteps","date")
```
The number of missing values (i.e., rows with NA for the step value) is `r numNA`, which corresponds to `r daysNA` days of missing values. The missing values (NAs) were filled in using the mean number of steps for that interval across all days, rounded to the nearest integer. A histogram of the total number of steps taken per day with the filled in data is shown below.  

```{r missingValueResults, fig.height=4}
# create ggplot histogram of total number of steps taken per day
stepHisto2 <- ggplot(totalStepsFill,aes(x=totSteps)) + geom_histogram(binwidth=1000,color="black",fill="dodgerblue4") + labs(x="Total Steps Taken Per Day (Filled in Data)",y="Frequency")
print(stepHisto2)
# calculate histogram data for the original data set (Na values removed)
stepHistoData <- hist(totalSteps$totSteps,plot=FALSE,breaks=25)
# determine the maximum frequency value for steps taken per day (i.e., count value)
maxFreqIdx <- which.max(stepHistoData$counts)
maxFreq <- stepHistoData$counts[maxFreqIdx]
# find the range of steps taken that corresponds to that maximum frequency value
maxLowRange <- stepHistoData$breaks[maxFreqIdx]
maxHighRange <- stepHistoData$breaks[maxFreqIdx+1]

# calculate histogram data for the filled in data set (Na values filled 
# with rounded mean # of steps)
stepHistoData2 <- hist(totalStepsFill$totSteps,plot=FALSE,breaks=25)
# similarly calculate maximum frequency value for steps taken per day
maxFreqIdx2 <- which.max(stepHistoData2$counts)
maxFreq2 <- stepHistoData2$counts[maxFreqIdx2]
# find the rnage of steps taken that corresponds to that maximum frequency 
# value for the filled in data
maxLowRange2 <- stepHistoData2$breaks[maxFreqIdx2]
maxHighRange2 <- stepHistoData2$breaks[maxFreqIdx2+1]

# prepare characters/strings to be displayed based on the differences in the
# step ranges for the maximum frequency values and the maximum frequency values
# for the original data set (NA values removed) and the inputted data set
# (NA values filled in)
totStepImpact <- "The inputting of rounded mean values for missing (NA) values "

if ((maxLowRange2 == maxLowRange) & (maxHighRange2 == maxHighRange)) {
    totStepImpact <- paste0(totStepImpact,"did not impact the most frequent total number of steps per day, ranging from ",maxLowRange," to ",maxHighRange,". ")
} else {
    totStepImpact <- paste0(totStepImpact,"impacted the most frequent total number of steps per day, originally from ",maxLowRange,"-",maxHighRange," to ",maxLowRange2,"-",maxHighRange2,". ")
}

if (maxFreq2 == maxFreq) {
    totStepImpact <- paste0(totStepImpact,"The inputting of rounded mean values for missing (NA) values did not impact the maximum frequency (count) of total number of steps per day.")
} else {
    totStepImpact <- paste0(totStepImpact,"The inputting of rounded mean values for missing (NA) values changed the maximum frequency (count) of total number of steps per day from ",maxFreq," to ",maxFreq2,".\n")
}
# calculate mean and median of total steps taken per day for the filled in dataset
meanTotStepsFill <- mean(totalStepsFill$totSteps,na.rm=TRUE)
meanTotStepsFill
medianTotStepsFill <- median(totalStepsFill$totSteps,na.rm=TRUE)
medianTotStepsFill
```
For the filled in data, the mean total number of steps taken per day is `r meanTotStepsFill` and the median total number of steps taken per day is `r medianTotStepsFill`. The difference in the mean total number of steps between the filled in data and the raw data is `r meanTotStepsFill - meanTotSteps` and the difference in median number of steps is `r medianTotStepsFill - medianTotSteps`. While the changes to the mean and the median number of steps taken per day is minimal (`r max((medianTotStepsFill - medianTotSteps)/medianTotSteps*100,(meanTotStepsFill - meanTotSteps)/meanTotSteps)*100`%), there was a significant change in the distribution of the total number of steps taken per day. `r totStepImpact`  

## Are there differences in activity patterns between weekdays and weekends?
```{r dayOfweekPattern}
# set day of week for each date the data was collected
activity.fill$dayOfWeek <- weekdays(activity.fill$date)
# set a factor variable for the daytype (either "weekday" or "weekend"), filled with NAs
activity.fill$dayType <- factor(rep(NA,nrow(activity.fill)),levels=c("weekday","weekend"))
# set dayType equal to "weekend" if dayOfWeek is "Saturday" or "Sunday"
activity.fill$dayType[activity.fill$dayOfWeek %in% c("Saturday","Sunday")] <- "weekend"
# set dayType equal to "weekday" if dayType was still an NA value
activity.fill$dayType[is.na(activity.fill$dayType)] <- "weekday"

pattern2 <- ggplot(activity.fill,aes(x=interval,y=steps, color=dayType)) + stat_summary(fun.y=mean,geom="line") + facet_wrap(~dayType, nrow=2) + labs(x="Time Interval",y="Mean Steps",title="Mean Steps Taken as a Function of Time of Day")
print(pattern2)
```
  
There are differences in the steps taken activity patterns between weekdays and weekends, as shown in the plot above. During the weekdays, on average more steps are taken earlier in the day (approximately 0530 to 0730). On the weekends, on average more steps are taken between 2000 to 2100.